---
title: "Workshop"
author: "Leslie Fischer"
date: "23/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# 1. Preparation 

Clear the workspace
```{r}
rm(list = ls()) 
```

Set the working directory to the folder containing the workshop files:
```{r}
setwd("~/Desktop/Workshop")
```

Install.packages(x) 
```{r}
x <- c("readxl", "sf", "igraph")
```

Load the required packages 
```{r results='hide', message=FALSE, warning=FALSE}
lapply(x, library, character.only = TRUE) 
```

Load data
```{r}
library(readxl)
nodes <- read_excel("NODES.xlsx")
links <- read_excel("EDGES.xlsx")
```

Pie chart 
```{r}

pie(nodes$visits, labels=nodes$country, col = rainbow(length(nodes$visits)), main= "Visits")
```


# 2. Spatial Data 
Adapted from 

1. Weidmann (forthcoming): Data Management for Social Scientists

2. Ward & Gleditsch (2008): Spatial Regression Models

Right now, R treats our data as a regular data frame, which is why we convert it into a spatial object. "coords=c(longitude, latitude)" requires us to specify the names of the columns where the spatial coordinates are stored.

Load sf package
```{r}
library(sf)
```

Enable dataset
```{r}
capital <- st_as_sf(nodes, coords = c("longitude_cap", "latitude_cap"), crs = 4326)
```

Check
```{r}
print(capital, n=4)
```

Load spatial data set for Europe
```{r}
europe <- st_read(file.path("NUTS_RG_60M_2021_4326"), crs=4326)
```

Plot spatial data set and add coordinates of capitals
```{r}
plot(st_geometry(europe), col="lightgrey")
plot(st_geometry(capital), pch=16, col = "red", add=T)
```


Too large: subset only the countries we are interested in: DE, IT, SE, PL
```{r}
better <- europe[europe$CNTR_CODE=="DE" | europe$CNTR_CODE=="IT" |
                 europe$CNTR_CODE=="PL"| europe$CNTR_CODE=="SE",]
```

Plot again
```{r}
plot(st_geometry(better), col="lightgrey")
plot(st_geometry(capital), pch=16, col = "red", add=T)
```

Now we want to color countries by number of visits. We therefore merge our data set with a data set which includes the information on the number of visits.
```{r}
better<- merge(better, nodes[,c("visits", "CNTR_CODE")], by = "CNTR_CODE", all = TRUE) 
```

How many features are in each category or level?
```{r}
summary(better$visits)
```

Convert to factor
```{r}
better$visits <- as.factor(better$visits)
class(better$visits)
```

Count the number of unique values or levels
```{r}
length(levels(better$visits))
```

Create a color palette of 6 colors - one for each factor level
```{r}
road_palette <- c("grey80", "grey60", "grey40", "grey20" )
road_palette
```

Create a vector of colors - one for each feature in the vector object according to its attribute value
```{r}
all_road_colors <- c("grey80", "grey60", "grey40", "grey20")[better$visits]
head(all_road_colors)
```

Plot the lines data, apply a different color to each factor level. Then add the coordinates of the capitals. When adding the legend, x="right" specifies the location of the legend. cex is the font size. After "fill=" we name the color palette we used to fill our objects with.
```{r}
plot(st_geometry(better), col= all_road_colors)
plot(st_geometry(capital), pch=16, col = "red", add=T)
legend(x = "right", c("  6 visits", "10 visits", "13 visits", "19 visits" ), cex = .5, fill = road_palette)
```


# 3. Network Data 
Adapted from: Ognyanova (2016): https://kateto.net/wp-content/uploads/2016/01/NetSciX_2016_Workshop.pdf

Load package
```{r}
library(igraph)
```

Enable data set
```{r}
net <- graph_from_data_frame(d=links, vertices=nodes, directed=T) 
```

# Plot network

edge.arrow.size=.4: size of arrow

vertex.color="lightgrey": node color 

vertex.label=V(net)$id: specifies label of node

vertex.label.font=2: font type for the label (1: Times new Roman, 2: bold, 3: italic+, 4: italic+bold)

vertex.label.color="black": color of label

vertex.size=40: size of node

vertex.label.cex=.7: size of label within node

edge.color="gray85": edge color

```{r}
plot(net, edge.arrow.size=.4, vertex.color="lightgrey", vertex.label=V(net)$id, 
     vertex.label.font=2, vertex.label.color="black", vertex.size=50,
     vertex.label.cex=.7, edge.color="gray85")
```

Set node (/vertix) size based on number of visits
```{r}
V(net)$size <- V(net)$visits*3
```

Plot (take out vertex.size=40)
```{r}
plot(net, edge.arrow.size=.4, vertex.color="lightgrey", vertex.label=V(net)$id, 
     vertex.label.font=2, vertex.label.color="black", 
     vertex.label.cex=.5, edge.color="gray85")
```


## BONUS: 
Color by long distance (>1,000 km)
```{r}
colrs <- E(net)$color <- ifelse( E(net)$dummy == "1", "tomato", "gold")
```

Plot and add legend
```{r}
plot(net, edge.arrow.size=.4, vertex.color="lightgrey", vertex.label=V(net)$id, 
     vertex.label.font=2, vertex.label.color="gray40", 
     vertex.label.cex=.5)
legend(x=-1.1, y=-1.5, c("Long-distance"), pch=21,
       col="#777777", pt.bg=colrs, pt.cex=2.5, bty="n", ncol=1)
```
